### 2025-01-16 sections/1-introduction.tex
yet more edits

Hierarchical Navigable Small World networks~\cite{malkov2016hnsw} relies on navigable small world (NSW) networks~\cite{kleinberg2000navigation, boguna2009navigability} and skip lists.
HNSW builds a multi-layered graph of the dataset.
A query point and each data point are inserted into the graph and joined by an edge to the $M$ nearest nodes in the in the graph, where $M$ is a tunable parameter.
The highest layer in which an element can be placed is determined randomly with an exponentially-decaying probability distribution.
Search starts at the highest layer and descends to the lowest layer, greedily following a path of edges to the nearest node, until reaching the query point.
To improve accuracy, the $efSearch$ hyperparameter can be changed to specify the number of closest nearest neighbors to the query vector to be found at each layer.

>>>>

Hierarchical Navigable Small World networks~\cite{malkov2016hnsw} rely on navigable small world (NSW) networks~\cite{kleinberg2000navigation, boguna2009navigability} and skip lists.
HNSW builds a multi-layered graph of the dataset.
A query point and each data point are inserted into the graph and joined by an edge to the $M$ nearest nodes in the in the graph, where $M$ is a tunable parameter.
The highest layer in which an element can be placed is determined randomly with an exponentially-decaying probability distribution.
Search starts at the highest layer and descends to the lowest layer, greedily following a path of edges to the nearest node, until reaching the query point.
To improve accuracy, the $efSearch$ hyperparameter can be changed to specify the number of closest nearest neighbors to the query vector to be found at each layer.

### 2025-01-16 sections/2-methods.tex
minor tweaks
yet more edits

We define a \textit{distance function} $f : \textbf{X} \times \textbf{X} \mapsto \mathbb{R}^+ \ \cup \ \{0\}$ which, given two points, deterministically returns a finite and non-negative real number.
A distance of zero defines identity among points (i.e.,\,$f(x, y) = 0 \Leftrightarrow x = y$) and larger values indicate greater dissimilarity among points.
We also require that the distance function be symmetric (i.e.,\,$f(x, y) = f(y, x) \ \forall \ x, y \in \textbf{X}$).
In addition to these constraints, if the distance function obeys the triangle inequality (i.e.,\,$f(x, y) \leq f(x, z) + f(z, y) \ \forall \ x, y, z \in \textbf{X}$), then it is also a \textit{distance metric}.
Similar to~\cite{yu2015entropy}, when used with distance metrics, all search algorithms in CAKES are exact.
For example, Euclidean, Levenshtein~\cite{levenshtein1966binary} and Dynamic Time Warping (DTW)~\cite{muller2007dynamic} distances are all distance metrics, while cosine distance is not a metric because it violates the triangle inequality (e.g.,\,consider the points $x = (1, 0)$, $y = (0, 1)$ and $z = (1, 1)$ on the Cartesian plane).

The choice of an appropriate distance function varies by dataset and domain.
For example, with neural-network embeddings, one could use Euclidean (L2) or cosine distance.
With genomic or proteomic sequence data, Levenshtein or Hamming distance is useful.
With time-series data, one could use Dynamic Time Warping (DTW) or Wasserstein distance.

@@

Intuitively, LFD measures the rate of change in the number of points in a ball of radius $r$ around a point $q$ as $r$ increases. When the vast majority of points in the dataset have low ($\ll D$) LFD, we can simply say that the dataset has low LFD.
We stress that this concept differs from the \textit{embedding dimension} of a dataset.
To illustrate the difference, consider the SILVA 18S rRNA dataset that contains genomic sequences with unaligned lengths of up to 3,712 base pairs and aligned length of 50,000 base pairs.
Hence, the \textit{embedding dimension} of this dataset is at least 3,712 and at most 50,000; as used in this paper it is 3,712.
However, physical constraints (namely, biological evolution and biochemistry) constrain the data to a lower-dimensional manifold within this space.
LFD is an approximation of the dimensionality of that lower-dimensional manifold in the ``vicinity'' of a given point.
Figure~\ref{fig:results:lfd-plots} illustrates this concept on a variety of datasets, showing how real datasets uphold the manifold hypothesis.
For real-world datasets, we expect the LFD to be locally uniform, i.e.,\,when $r$ is small, but potentially highly variable at larger length scales, i.e.,\,when $r$ is large.

@@

\begin{algorithm} % enter the algorithm environment
    \caption{Partition($C$, $criteria$)} % give the algorithm a caption
    \label{alg:methods:partition} % and a label for \ref{} commands later in the document
    \begin{algorithmic} % enter the algorithmic environment
        \Require $C$, a cluster
        \Require $criteria$, user-specified continuation criteria

@@

\subsubsection{Complexity}
\label{sec:methods:clustering:complexity}

@@

When building a tree for a real dataset, the depth would be larger than $\log n$, which would seem to increase the cost of tree-building.
However, because of the large numbers of leaf clusters at shallow depths, each subsequent level of $\mathcal{T}$ would also have fewer points, and so the cost of building each subsequent level would be lower.
Any analysis of unbalanced trees would be highly dependent on the specific dataset, and so we do not provide one here, but we explore empirical results based on forcing a balanced tree in Section~\ref{sec:results:clustering-strategies-and-number-of-distance-computations}.
In any case, the tree is built only once, so this cost is amortized over all search queries and other applications.

@@

\begin{figure}[H]
    \centering
    \includegraphics[scale=0.65]{images/geometry/deltas.pdf}
    \caption{
        {\color{blue}$\delta$}, {\color{red}$\delta^{+}$}, and {\color{green}$\delta^{-}$} for a cluster $C$ and a query $q$.
        ${\color{blue}\delta} = f(q, c)$ is the distance from the query to the cluster center $c$.
        ${\color{red}\delta^{+}} = \delta + r$ is the distance from the query to the theoretically farthest point in $C$.
        ${\color{green}\delta^{-}} = \text{max}(0, \delta - r)$ is the distance from the query to the theoretically closest point in $C$.
    }
    \label{fig:methods:deltas}
    \vskip -0.25in
\end{figure}

\begin{minipage}{.425\textwidth}
    \begin{algorithm}[H]
        \caption{tree-search($C$, $q$, $r$)}
        \label{alg:methods:rnn-search:tree-search}
        \begin{algorithmic}
            \Require $C$, a cluster
            \Require $q$, a query
            \Require $r$, a search radius
            \If{$\delta^+_C \leq r$}
                \State \textbf{return} $\{C\}$
            \Else
                \State $[L, R]$ $\Leftarrow$ \textit{children} of $C$
                \State \textbf{return} tree-search($L, q, r$) \\
                \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ $\cup$ tree-search($R, q, r$)
            \EndIf
        \end{algorithmic}
    \end{algorithm}
\end{minipage}
\hfill
\begin{minipage}{.475\textwidth}
    \begin{algorithm}[H]
        \caption{leaf-search($Q$, $q$, $r$)}
        \label{alg:methods:rnn-search:leaf-search}
        \begin{algorithmic}
            \Require $Q$, a set of clusters
            \Require $q$, a query
            \Require $r$, a search radius
            \State $H \Leftarrow \emptyset$
            \For{$C \in Q$}
                \If{$\delta^+_C \leq r$}
                    \State $H$ $\Leftarrow$ $H \cup C$
                \Else
                    \For{$p \in C$}
                        \If{$f(p, q) \leq r$}
                            \State $H$ $\Leftarrow$ $H \cup \{p\}$
                        \EndIf
                    \EndFor
                \EndIf
            \EndFor
            \State \textbf{return} $H$
        \end{algorithmic}
    \end{algorithm}
\end{minipage}

@@

As stated in Section~\ref{sec:intoduction:entropy-scaling-search}, Theorem~\ref{thm:methods:rnn-complexity} assumes that the dataset $X$ has a manifold structure.

% The complexity bounds for Repeated $\rho$-NN rely on the assumption that the query point is sampled from the same distribution as the rest of the data or, in other words, that it arises from the same generative process as the rest of the dataset.
% Given the uses of $k$-NN search in practice, this assumption is reasonable.
% From this assumption, we can infer that the LFD near the query does not differ significantly from the (harmonic) mean of the LFDs of clusters near the query at the scale of the distance from the query to the $k^{th}$ nearest neighbor.

\begin{theorem} Let $X$ be a dataset and $q$ a query sampled from the same distribution (i.e., arising from the same generative process) as $X$. Then time complexity of performing Repeated $\rho$-NN search on $X$ with query $q$ is \begin{gather}
        \mathcal{O}
        \Bigg(
            \underbrace{
                \log~\overbrace{\mathcal{N}_{\hat{r}}(X)}^{\textrm{metric entropy}}
            }_{\textrm{tree-search}}
            \ + \
            \underbrace{
                \overbrace{k}^{\textrm{output size}} \cdot
                \overbrace{\bigg( 1 + 2 \cdot \Big( \frac{\hat{|C|}}{k} \Big) ^ {d^{-1}} \bigg)^d}^{\textrm{scaling factor}}
            }_{\textrm{leaf-search}}
        \Bigg)
        \label{eq:methods:repeated-rnn-complexity1}
    \end{gather}
    where $\mathcal{N}_{\hat{r}}(X)$ is the metric entropy of the dataset, $d$ is the LFD of the dataset, and $k$ is the number of nearest neighbors.
    \label{thm:methods:rnn-complexity}
\end{theorem}

% \textbf{Theorem.} \textit{Let $X$ be a dataset and $q$ a query sampled from the same distribution (i.e., arising from the same generative process) as $X$. Then time complexity of performing Repeated $\rho$-NN search on $X$ with query $q$ is} \begin{gather}
%     \mathcal{O}
%     \Bigg(
%         \underbrace{
%             \log~\overbrace{\mathcal{N}_{\hat{r}}(X)}^{\textrm{metric entropy}}
%         }_{\textrm{tree-search}}
%         \ + \
%         \underbrace{
%             \overbrace{k}^{\textrm{output size}} \cdot
%             \overbrace{\bigg( 1 + 2 \cdot \Big( \frac{\hat{|C|}}{k} \Big) ^ {d^{-1}} \bigg)^d}^{\textrm{scaling factor}}
%         }_{\textrm{leaf-search}}
%     \Bigg).
%     \label{eq:methods:repeated-rnn-complexity1}
% \end{gather}

@@

The work in~\cite{yu2015entropy} showed that $\sum_{C \in S} |C| \leq \gamma  \left| B(q, \rho_k) \right| \left(\frac{\rho_k + 2 \cdot \hat{r}}{\rho_k} \right)^d$, where $\gamma$ is a constant.
By definition of $\rho_k$, we have that $|B(q, \rho_k)| = k$.
Thus, $\sum_{C \in S} |C| \leq \gamma k \left( 1 + 2 \cdot \frac{\hat{r}}{\rho_k} \right)^d$.
It remains to determine an estimate for $\rho_k$.
% For this, we once again rely on the assumption that the query is drawn from the same distribution as the rest of the data, and thus the LFD at the query point is not significantly different from the LFD of nearby clusters.
We let $\hat{d}$ be the (harmonic) mean LFD of the clusters in $Q$.
While ordinarily we compute LFD by comparing cardinalities of two balls with two different radii centered at \textit{the same} point, in order to estimate $\rho_k$, we instead compare the cardinality of a ball \textit{around the query} of radius $\rho_k$ to the mean cardinality, $\hat{|C|}$, of clusters in $Q$ at a radius equal to the mean of their radii, $\hat{r}$.
Since $q$ is from the same distribution as $X$, the LFD at $q$ should not be significantly different from that at the center of a cluster in $Q$.
By Equation~\ref{eq:methods:lfd-original}, $\hat{d} = \frac{\log{}\frac{\hat{|C|}}{k}}{\log{}\frac{\hat{r}}{\rho_k}}$.
We can rearrange this equation to get $\frac{\hat{r}}{\rho_k} = \left( \frac{\hat{|C|}}{k} \right)^{\hat{d}^{-1}}$.
Using this to simplify the term for leaf-search in Equation~\ref{eq:methods:rnn-search-complexity}, we get:
\begin{equation*}
    k \left( 1 + 2 \cdot \left( \frac{\hat{|C|}}{k} \right) ^ {\hat{d}^{-1}} \right)^d
\end{equation*}

Again using the assumption that since $q$ is from the same distribution as $X$, the LFD at $q$ should not be significantly different from that at a cluster in $Q$, we have that $\hat{d} \approx d$.
By combining the bounds for tree-search and leaf-search, we see that Repeated $\rho$-NN has time complexity claimed in Equation~\ref{eq:methods:repeated-rnn-complexity1}.
\end{proof}

We remark that the scaling factor in Equation~\ref{eq:methods:repeated-rnn-complexity1} should be close to 1 unless LFD is highly variable in the region around the query (i.e.,\,if $\hat{d}$ differs significantly from $d$).

@@

Due to their similarity, we combine the complexity analyses of both Sieve methods.
For these methods we again use the terminology of tree-search and leaf-search.
Tree-search navigates the cluster tree and adds clusters to $Q$.
Leaf-search exhaustively searches \textit{some} of the clusters in $Q$ to find the $k$ nearest neighbors.
As stated in Section~\ref{sec:intoduction:entropy-scaling-search}, Theorem~\ref{thm:methods:sieve-complexity} assumes that the dataset $X$ has a manifold structure.

@@

The asymptotic complexity of Breadth-First Sieve is dominated by the QuickSelect algorithm to calculate $\tau$.
Since this method is log-linear in the length of $Q$, and $Q$ contains the clusters from tree-search and the points from leaf-search, we see that the asymptotic complexity is
\begin{equation*}
    \mathcal{O} \Big( (T + L ) \log (T + L ) \Big),
\end{equation*} as claimed.

For Depth-First Sieve, since we use two priority queues, the asymptotic complexity is dominated by the priority queue operations.
Thus, the complexity is
\begin{equation*}
    \mathcal{O} \Big( T \log T + L \log k \Big),
\end{equation*} as claimed.
\end{proof}

>>>>

We define a \textit{distance function} $f : \textbf{X} \times \textbf{X} \mapsto \mathbb{R}^+ \ \cup \ \{0\}$ which, given two points, deterministically returns a finite and non-negative real number.
A distance value of zero defines an identity among points (i.e.,\,$f(x, y) = 0 \Leftrightarrow x = y$) and larger values indicate greater dissimilarity among points.
We also require that the distance function be symmetric (i.e.,\,$f(x, y) = f(y, x) \ \forall \ x, y \in \textbf{X}$).
In addition to these constraints, if the distance function obeys the triangle inequality (i.e.,\,$f(x, y) \leq f(x, z) + f(z, y) \ \forall \ x, y, z \in \textbf{X}$), then it is also a \textit{distance metric}.
Similar to~\cite{yu2015entropy}, when used with distance metrics, all search algorithms in CAKES are exact.
For example, Euclidean, Levenshtein~\cite{levenshtein1966binary} and Dynamic Time Warping (DTW)~\cite{muller2007dynamic} distances are all distance metrics, while cosine distance is not a metric because it violates the triangle inequality (e.g.,\,consider the points $x = (1, 0)$, $y = (0, 1)$ and $z = (1, 1)$ on the Cartesian plane).

The choice of an appropriate distance function varies by dataset and domain.
For example, with neural-network embeddings, one could use Euclidean or Cosine distance.
With genomic or proteomic sequence data, Levenshtein and Hamming distances are useful.
With time-series data, one could use Dynamic Time Warping (DTW) or Wasserstein distance.

@@

Intuitively, LFD measures the rate of change in the number of points in a ball of radius $r$ around a point $q$ as $r$ increases. When the vast majority of points in the dataset have low ($\ll D$) LFD, we can simply say that the dataset has low LFD.
We stress that this concept differs from the \textit{embedding dimension} of a dataset.
To illustrate the difference, consider the SILVA 18S rRNA dataset that contains genomic sequences with unaligned lengths of up to 3,712 base pairs and aligned length of 50,000 base pairs.
Hence, the \textit{embedding dimension} of this dataset is at least 3,712 and at most 50,000; as used in this paper it is 3,712.
However, physical constraints (namely, biological evolution and biochemistry) constrain the data to a lower-dimensional manifold within this space.
LFD is an approximation of the dimensionality of that lower-dimensional manifold in the ``vicinity'' of a given point.
Section~\ref{sec:results:lfd-of-datasets} discusses this concept on a variety of datasets, showing how real datasets uphold the manifold hypothesis.
For real-world datasets, we expect the LFD to be locally uniform, i.e.,\,when $r$ is small, but potentially highly variable at larger length scales, i.e.,\,when $r$ is large.

@@

\begin{algorithm} % enter the algorithm environment
    \caption{Partition($C$, $criteria$)} % give the algorithm a caption
    \label{alg:methods:partition} % and a label for \ref{} commands later in the document
    \begin{algorithmic} % enter the algorithmic environment
        \Require $f: X \times X \mapsto \mathbb{R}^+ \cup \{0\}$, a distance function
        \Require $C$, a cluster
        \Require $criteria$, user-specified continuation criteria

@@

\subsubsection{Time Complexity}
\label{sec:methods:clustering:time-complexity}

@@

When building a tree for a real dataset, the depth would be larger than $\log n$, which would seem to increase the cost of tree-building.
However, because of the large numbers of leaf clusters at shallow depths, each subsequent level of $\mathcal{T}$ would also have fewer points, and so the cost of building each subsequent level would be lower.
Any analysis of unbalanced trees would be highly dependent on the specific dataset, and so we do not provide one here.
We do, however, explore empirical results based on forcing a balanced tree in Section~\ref{sec:results:clustering-strategies-and-number-of-distance-computations}.
In any case, the tree is built only once, so this cost is amortized over all search queries and other applications.

\begin{figure}[H]
    \vskip -0.2in
    \centering
    \includegraphics[scale=0.65]{images/geometry/deltas.pdf}
    \caption{
        {\color{blue}$\delta$}, {\color{red}$\delta^{+}$}, and {\color{green}$\delta^{-}$} for a cluster $C$ and a query $q$.
        ${\color{blue}\delta} = f(q, c)$ is the distance from the query to the cluster center $c$.
        ${\color{red}\delta^{+}} = \delta + r$ is the distance from the query to the theoretically farthest point in $C$.
        ${\color{green}\delta^{-}} = \text{max}(0, \delta - r)$ is the distance from the query to the theoretically closest point in $C$.
    }
    \label{fig:methods:deltas}
    \vskip -0.5in
\end{figure}

\begin{minipage}{.425\textwidth}
    \begin{algorithm}[H]
        \caption{tree-search($C$, $q$, $r$)}
        \label{alg:methods:rnn-search:tree-search}
        \begin{algorithmic}
            \Require $f$, a distance function
            \Require $C$, a cluster
            \Require $q$, a query
            \Require $r$, a search radius
            \If{$\delta^+_C \leq r$}
                \State \textbf{return} $\{C\}$
            \Else
                \State $[L, R]$ $\Leftarrow$ \textit{children} of $C$
                \State \textbf{return} tree-search($L, q, r$) \\
                \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ $\cup$ tree-search($R, q, r$)
            \EndIf
        \end{algorithmic}
    \end{algorithm}
\end{minipage}
\hfill
\begin{minipage}{.475\textwidth}
    \begin{algorithm}[H]
        \caption{leaf-search($Q$, $q$, $r$)}
        \label{alg:methods:rnn-search:leaf-search}
        \begin{algorithmic}
            \Require $f$, a distance function
            \Require $Q$, a set of clusters
            \Require $q$, a query
            \Require $r$, a search radius
            \State $H \Leftarrow \emptyset$
            \For{$C \in Q$}
                \If{$\delta^+_C \leq r$}
                    \State $H$ $\Leftarrow$ $H \cup C$
                \Else
                    \For{$p \in C$}
                        \If{$f(p, q) \leq r$}
                            \State $H$ $\Leftarrow$ $H \cup \{p\}$
                        \EndIf
                    \EndFor
                \EndIf
            \EndFor
            \State \textbf{return} $H$
        \end{algorithmic}
    \end{algorithm}
\end{minipage}

@@

@@

\begin{theorem} Let $X$ be a dataset and $q$ a query sampled from the same distribution (i.e., arising from the same generative process) as $X$. Then time complexity of performing Repeated $\rho$-NN search on $X$ with query $q$ is \begin{gather}
        \mathcal{O}
        \Bigg(
            \underbrace{
                \log~\overbrace{\mathcal{N}_{\hat{r}}(X)}^{\textrm{metric entropy}}
            }_{\textrm{tree-search}}
            \ + \
            \underbrace{
                \overbrace{k}^{\textrm{output size}} \cdot
                \overbrace{\bigg( 1 + 2 \cdot \Big( \frac{\hat{|C|}}{k} \Big) ^ {d^{-1}} \bigg)^d}^{\textrm{scaling factor}}
            }_{\textrm{leaf-search}}
        \Bigg)
        \label{eq:methods:repeated-rnn-complexity}
    \end{gather}
    where $\mathcal{N}_{\hat{r}}(X)$ is the metric entropy of the dataset, $d$ is the LFD of the dataset, and $k$ is the number of nearest neighbors.
    \label{thm:methods:rnn-complexity}
\end{theorem}

@@

The work in~\cite{yu2015entropy} showed that $\sum_{C \in S} |C| \leq \gamma  \left| B(q, \rho_k) \right| \left(\frac{\rho_k + 2 \cdot \hat{r}}{\rho_k} \right)^d$, where $\gamma$ is a constant.
By definition of $\rho_k$, we have that $|B(q, \rho_k)| = k$.
Thus, $\sum_{C \in S} |C| \leq \gamma k \left( 1 + 2 \cdot \frac{\hat{r}}{\rho_k} \right)^d$.
It remains to determine an estimate for $\rho_k$.
We let $\hat{d}$ be the (harmonic) mean LFD of the clusters in $Q$.
While ordinarily we compute LFD by comparing cardinalities of two balls with two different radii centered at \textit{the same} point, in order to estimate $\rho_k$, we instead compare the cardinality of a ball \textit{around the query} of radius $\rho_k$ to the mean cardinality, $\hat{|C|}$, of clusters in $Q$ at a radius equal to the mean of their radii, $\hat{r}$.
Since $q$ is from the same distribution as $X$, the LFD at $q$ should not be significantly different from that at the center of a cluster in $Q$.
By Equation~\ref{eq:methods:lfd-original}, $\hat{d} = \frac{\log{}\frac{\hat{|C|}}{k}}{\log{}\frac{\hat{r}}{\rho_k}}$.
We can rearrange this equation to get $\frac{\hat{r}}{\rho_k} = \left( \frac{\hat{|C|}}{k} \right)^{\hat{d}^{-1}}$.
Using this to simplify the term for leaf-search in Equation~\ref{eq:methods:rnn-search-complexity}, we get:
\begin{equation*}
    k \left( 1 + 2 \cdot \left( \frac{\hat{|C|}}{k} \right) ^ {\hat{d}^{-1}} \right)^d
\end{equation*}

Again using the assumption that since $q$ is from the same distribution as $X$, the LFD at $q$ should not be significantly different from that at a cluster in $Q$, we have that $\hat{d} \approx d$.
By combining the bounds for tree-search and leaf-search, we see that Repeated $\rho$-NN has time complexity as claimed in Equation~\ref{eq:methods:repeated-rnn-complexity}.
\end{proof}

We remark that the scaling factor in Equation~\ref{eq:methods:repeated-rnn-complexity} should be close to 1 unless LFD is highly variable in the region around the query (i.e.,\,if $\hat{d}$ differs significantly from $d$).

@@

Due to their similarity, we combine the complexity analyses of both Sieve methods.
For these methods we again use the terminology of tree-search and leaf-search.
Tree-search navigates the cluster tree and adds clusters to $Q$.
Leaf-search exhaustively searches \textit{some} of the clusters in $Q$ to find the $k$ nearest neighbors.

@@

The asymptotic complexity of Breadth-First Sieve is dominated by the QuickSelect algorithm to calculate $\tau$.
Since this method is log-linear in the length of $Q$, and $Q$ contains the clusters from tree-search and the points from leaf-search, we see that the asymptotic complexity is as claimed in Equation~\ref{eq:methods:breadth-first-sieve-complexity}.

For Depth-First Sieve, since we use two priority queues, the asymptotic complexity is as claimed in Equation~\ref{eq:methods:depth-first-sieve-complexity}.
\end{proof}

### 2025-01-16 sections/3-datasets.tex
yet more edits

To demonstrate CAKES with a more exotic distance function, we also benchmarked on the SILVA 18S ribosomal RNA dataset~\cite{10.1093/nar/gks1219}.
This dataset contains ribosomal RNA sequences of 2,224,640 genomes, with the longest sequence having 3,712 letters.
We use Levenshtein~\cite{levenshtein1966binary} distance on the unaligned sequences to build the tree and to perform $k$-NN search.
We note that when put into a multiple sequence alignment, the length (and thus dimensionality) would be 50,000, but in that case one could use Hamming distance, which is computationally less expensive than Levenshtein distance.
We held out a set of 1,000 random sequences from the dataset to use as queries for benchmarking.
Since this dataset contains many redundant, i.e.,\,nearly identical, sequences, we used a maximum tree-depth of 128 as a partition criterion in Algorithm~\ref{alg:methods:partition}.

@@

As another example of an exotic distance function, we benchmarked CAKES on the RadioML dataset~\cite{oshea2018radioml}.
The RadioML dataset contains samples of synthetically generated signal captures of different modulation modes over a range of SNR levels.
Specifically, it comprises 24 modulation modes at 26 different signal-to-noise ratio (SNR) levels ranging from -20 dB to 30 dB, with 4,096 samples at each combination of modulation mode and SNR level.
Thus, it contains $24 \cdot 26 \cdot 4096 = 2,555,504$ samples in total.
Each sample is a 1,024-dimensional complex-valued vector, representing a signal capture (a time-series of complex-valued numbers).
We used a subset of this dataset, containing $97,304$ samples at 10dB SNR and used another $1,000$ samples at the same SNR as a hold-out set of queries.
We use Dynamic Time Warping~\cite{muller2007dynamic} as the distance function for this dataset.

>>>>

To demonstrate CAKES with a more exotic distance function, we also benchmarked on the SILVA 18S ribosomal RNA dataset~\cite{10.1093/nar/gks1219}.
This dataset contains ribosomal RNA sequences of 2,224,640 genomes, with the longest sequence having 3,712 letters.
We held out a set of 1,000 random sequences from the dataset to use as queries for benchmarking.
We use Levenshtein~\cite{levenshtein1966binary} distance on the unaligned sequences to build the tree and to perform $k$-NN search.
We note that the sequences in this dataset were provided in a multiple sequence alignment with a width of 50,000 characters.
We could have used Hamming distance on the aligned sequences, but we chose to use Levenshtein distance on the unaligned sequences to help demonstrate the flexibility of CAKES in handling exotic distance functions.
Under hamming distance, we could consider this dataset to have an embedding dimension of 50,000, but for Levenshtein distance, we state the dimensionality as the length of the longest sequence, i.e.\, 3,712.

@@

As another example of an exotic distance function, we benchmarked CAKES on the Radio-ML dataset~\cite{oshea2018radioml} under Dynamic Time Warping~\cite{muller2007dynamic} as the distance function.
This dataset contains samples of synthetically generated signal captures of different modulation modes over a range of signal-to-noise ratio (SNR) levels.
Specifically, it comprises 24 modulation modes at 26 different SNRs ranging from -20 dB to 30 dB, with 4,096 samples at each combination of modulation mode and SNR level.
Thus, it contains $24 \cdot 26 \cdot 4096 = 2,555,504$ samples in total.
Each sample is a 1,024-dimensional complex-valued vector, representing a signal capture (a time-series of complex-valued numbers).
We used a subset of this dataset, containing $97,304$ samples at 10dB SNR and used another $1,000$ samples at the same SNR as a hold-out set of queries.
